\chapter{Introduction}
\label{cha:introduction}

\begin{quote}
   This chapter introduces the pylinear Python extension and outlines the rest
   of the document.
\end{quote}

PyLinear is a set of extensions to the Python programming language
which allows Python programmers to efficiently manipulate matrices and
vectors, the primary objects of linear algebra.  It allows real and
complex arithmetic, currently only in double precision.  Dense as well
as two types of sparse matrices are supplied, and a large variety of
numerical algorithms, from eigensolvers, to singular value
decomposition, direct sparse solvers to sparse eigensolvers are also
furnished as part of an ever-growing standard library.

PyLinear's programming interface is similar to that of Numerical
Python and numarray to ease porting, but differs in a few key
aspects. The most notable such aspect is matrix multiplication. While
the term \code{A*B} in Numeric means element-by-element
multiplication, PyLinear changes this to mean conventional
matrix-matrix and matrix-vector multiplication, to match customary
uses in scientific computing, and following the example of languages
such as Matlab.  Chapter \ref{cha:diff-pylinear-numpy} is dedicated to
highlighting the differences between PyLinear and NumPy and its
desecendents.

In very simple terms, PyLinear is no more than a mapping of the
operators supplied by Boost.UBlas into Python using the Boost.Python
binding library.  This has two implications that balance each
other. First, PyLinear is no speed demon. It does have the right
asymptotic complexity guarantees (i.e. operations that ought to be
linear-time in fact are), but often, due to Python's interpreted
nature, the constants are pretty large. That's the bad news. The good
news is that since PyLinear is essentially a scripting language for
Boost.UBlas, it is sheepishly easy to convert a slow inner loop from
Python into C++, without losing much abstraction: The matrix and
vector types as well as most operations are available in C++ with only
slightly more difficult syntax than in Python. But if that is so, why
would you want to use Python in the first place? Because it's
high-level, safe and does not require the sometimes lengthy compile
times of C++.  And you need to convert \emph{only} that slow inner
loop!  Since you, too, can use Boost.Python to bind that inner loop to
Python (and still use PyLinear's facilities), there's no real need to
move the whole system into C++. That way, Python can be the convenient
and safe prototyping language for large computation systems written in
C++.
